Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2023-05-05 manuscript/manuscript.qmd
Mention WHO

Given recent medical advancements such as the distribution of vaccines globally and declining concern about COVID-19 severity, the Biden administration announced that it would end the public health emergency in the US on May 11, 2023. Doing so lifts measures instituted at the beginning of the pandemic—a move that has occurred in many other countries as well. As such, it is a worthwhile time to assess important descriptive and analytical questions about states’ respect for international law and human rights during this emergency period. This research note examines what derogating meant for compliance with international law during the COVID-19 pandemic. Specifically, we examine whether states actually derogated as per international legal principles and investigate the broader concern about whether states used derogations to exploit emergency provisions and to abuse human rights during COVID-19. These questions are important for the study of international law, compliance, and human rights during times of crises.

>>>>

Given recent medical advancements such as the distribution of vaccines globally and declining concern about COVID-19 severity, pandemic-related policies are less focused on rapid emergency responses. On May 5, 2023, the World Health Organization declared an end to the COVID-19 global health emergency, and many countries have made similar moves—the Biden administration, for instance, announced the end the United States' emergency measures on May 11, 2023. As such, now is a worthwhile time to assess important descriptive and analytical questions about states’ respect for international law and human rights during this emergency period. This research note examines what derogating meant for compliance with international law during the COVID-19 pandemic. Specifically, we examine whether states actually derogated as per international legal principles and investigate the broader concern about whether states used derogations to exploit emergency provisions and to abuse human rights during COVID-19. These questions are important for the study of international law, compliance, and human rights during times of crises.
